Extracting The Power of Dependent Types

### 2021-07-13 paper.tex
Brushed conclusions.

In this paper we investigate the idea of developing
embedded programs hand-in-hand with custom
code generators for them. We solve the well-known
conundrum of choosing between deep and shallow
embedding by leveraging the power of
reflection.
This allows us to enjoying the benefits
of shallow embedding, while keeping full access to the internal
structure of the embedded programs.

We apply this idea in the context of dependently-typed
languages to create verified implementations that can
be used in the context of an existing programming language.
We embed the target language in a theorem-prover, using
dependent types to encode the properties of interest.
Using reflection we bring the verified implementation
back into the original language.
%
Concretely, we have demonstrated the approach by implementing
an extractor for a fragment of the PostScript language, using
Agda as our host.

The main advantages of our approach are twofold.  First,
our solution is fine-grained --- we can chose what part
of the application to embed, and what constructs of the
host language to extract.  Secondly, our extractors
can use the full power of dependent types to guarantee
safety of our embedded programs.

\paragraph{Limitations}
\todo[inline]{This has moved from the intro, needs reflowing!}
While we claim our approach is flexible and powerful, it is
certainly not a magic bullet. There are still a lot of things that
need to be implemented for each target language: the language
embedding itself, the encoding of the properties under consideration
as types in the host language, and the extractor that transforms
embedded programs back into the syntax of the original language.  For
enforcing more complex properties, it might also be necessary to write
proofs to convince the typechecker of the host language that they are
indeed satisfied.
%
The main advantage of our approach is twofold: first, it allows
you to implement all these things --- as well as the embedded programs
themselves --- \emph{side by side in a single language}, simplifying
the development. And secondly, it allows you to reuse the typechecker
and evaluator of the host language, saving you a lot of work.  In the
end, our goal is not to showcase a single example but to empower you
to apply this approach yourself to verify code in any language.

\paragraph{Scaling to bigger languages}

The main goal of this paper is to demonstrate how to define a
embedding and extractor using dependent types and reflection on a
small but complete example. In addition to the PostScript embedding
discussed in this paper, we applied our approach to several other
languages:
\begin{itemize}
\item Kaleidoscope, a minimalist imperative language used in the
  documentation of LLVM.
\item SaC (Single assignment C), a high-performance array language
  that generates efficient code for various architectures.
\item APL (Array Programming Language), another array language with
  heavily overloaded syntax that pioneered the concept of rank- and
  shape-polymorphic programming.
\end{itemize}
In addition, we used the embedding of APL to implement a Convolutional
Neural Network, and ensure that all the array instructions operate on
arrays of the correct dimensions. For more details on these examples,
refer to \citet{DBLP:journals/corr/abs-2105-10819}.

\paragraph{Future work}

Right now we cannot yet guarantee that the
extracted code preservers the semantics of the original
implementation. While we rarely see
fully-verified compiler backends in the real world,
our approach is very close to enabling this.
We would need a formal semantics of the reflected language
and the proof that reflected programs respect it.
While this is non-trivial, a system like Agda could do
this in principle.

There is a number of improvements that can be added to
Agda and our extractors to make the resulting code more
efficient.  Supporting \AK{let}s in the internal syntax
would help to preserve sharing.  Recognising irrelevance
annotations in the extractors would help to eliminate
unused function arguments.  Introducing proper language
primitives to specify what exactly is an embedding would
be helpful. And finally, having access to more of
Agda's internals such as case trees would help to
generate more performant code.

Overall, this work only scratches the surface of extraction-based
compilation.  We never considered alternative theories supported
by theorem provers, \eg{} cubical type theory in Agda; we did not
consider recursive metaprogramming; we did not consider integrating
optimisations of extracted programs other than what rewriting rules
are capable to do.  All of these offer exciting research opportunities
on the way to make verified software easily accessible.
To paraphrase Jim Morrison: ``Safety and no surprise, The End''.

>>>>

In this paper we investigate the idea of developing
embedded programs hand-in-hand with custom
code generators for them. We solve the well-known
conundrum of choosing between deep and shallow
embedding by leveraging the power of
reflection.
This allows us to enjoy simplicity of writing progams
in shallow embedding.  Yet we keep full access to the internal
structure of the embedded programs, so that we could translate
them into the original language.

We apply this idea in the context of dependently-typed
language Agda.  We demonstrate the use of dependent types
in order to enforce static propeties of the embedded programs.
Concretely, we have demonstrated the approach by implementing
an extractor for a fragment of the PostScript language.

The main advantage of our approach is twofold: first, it allows
you to implement all these things --- as well as the embedded programs
themselves --- \emph{side by side in a single language}, simplifying
the development. And secondly, it allows you to reuse the typechecker
and evaluator of the host language, saving you a lot of work.
While we showcase a single example, we applied our approach to several other
languages:
\begin{itemize}
\item Kaleidoscope, a minimalist imperative language used in the
  documentation of LLVM.
\item SaC (Single assignment C), a high-performance array language
  that generates efficient code for various architectures.
\item APL (Array Programming Language), another array language with
  heavily overloaded syntax that pioneered the concept of rank- and
  shape-polymorphic programming.
\end{itemize}
For more details on these examples,
refer to~\citet{DBLP:journals/corr/abs-2105-10819}.
This sugests that you are now empowered to apply the proposed
technique tp verify code in any language.

\paragraph{Future work}
While our approach is flexible and powerful, it is
certainly not a magic bullet. A lot of things
need to be implemented for each target language: the embedding,
the encoding of the properties under consideration as types,
and the extractor.  For enforcing more complex properties, a number
of proofs needs to be written explicitly.

Right now, we cannot yet guarantee that the
extracted code preservers the semantics of the original
implementation. While we rarely see
fully-verified compiler backends in the real world,
our approach is very close to enabling this.
We would need a formal semantics of the reflected language
and the proof that reflected programs respect it.
While this is non-trivial, a system like Agda could do
this in principle.

% There is a number of improvements that can be added to
% Agda and our extractors to make the resulting code more
% efficient.  Supporting \AK{let}s in the internal syntax
% would help to preserve sharing.  Recognising irrelevance
% annotations in the extractors would help to eliminate
% unused function arguments.  Introducing proper language
% primitives to specify what exactly is an embedding would
% be helpful. And finally, having access to more of
% Agda's internals such as case trees would help to
% generate more performant code.
% 
% Overall, this work only scratches the surface of extraction-based
% compilation.  We never considered alternative theories supported
% by theorem provers, \eg{} cubical type theory in Agda; we did not
% consider recursive metaprogramming; we did not consider integrating
% optimisations of extracted programs other than what rewriting rules
% are capable to do.  All of these offer exciting research opportunities
% on the way to make verified software easily accessible.
% To paraphrase Jim Morrison: ``Safety and no surprise, The End''.
